\documentclass[letterpaper,12pt]{article}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{fancyheadings}
\usepackage{hyperref}
\pagestyle{fancy}
\usepackage{pmc}
\usepackage{graphicx}
\setlength\textwidth{6.5in}
\setlength\textheight{8.5in}
\input{../styles/ramesh_abbreviations}
\begin{document}
\title{An Internal MarketPlace}
\author{several authors}
\maketitle
\thispagestyle{fancy}
\lhead{}
\chead{}
\rhead{}
\lfoot{{\small Platform Team}}
\cfoot{}
\rfoot{{\small \thepage}}

\section{Introduction}

Let us start with a simple motivation. We have several high quality content
pieces in WordPress. On each of these, we would like to present ``related
content''. From our perspective, we hope to increase user engagement, which is
currently stubbornly low.  The choice of what content to present is dictated
by the following factors, listed in decreasing order of our ablity to execute
against them

\be
\item the content on the page vying for recommendation
\item the content on the main page 
\item information that come from the browser, ranging from
time of day to platform to geo-location
\item articles viewed in previous sessions, obtained from DWH
\item articles viewed in the current session
\ee

It is easy to generalize the notion of ``related'' beyond content as follows.
Let us consider a few examples
\be
\item A product manager would like to draw attention to their product e.g.,
  Elena wants more sign-ups for DIT
\item A general manager would like to meet business goals e.g., Kevin notices a
  potential shortfall on Citi Double Cash and would like to promote it
\item Strategic decision making e.g., Tim would like to experiment with some
  CTA's (Call To Action) 
  \ee

Notice that these are competing interests, that while working for the common
good, are locally optimizing for their specific needs.  The natural 
generalization of a solution to this problem is that of an auction house, where
the budgets for each bidder is set by the executive team. The act of setting
budgets is to make explicit the relative priorities of the competing interests.
While we will report on the efficacy of the placements, we leave it to each
bidder to determine what is good for them. If they choose to squander their
monthly budget on placing their ``ad'' where it has no chance of being clicked,
that is for them to worry about, not us. 
\end{document}
