\startreport{KeyCounter}
\reportauthor{Ramesh Subramonian}

\section{Introduction}
The KeyCounter is a hash table that is used to count the number of distinct
occurrences of a set of composite keys. We intend to extend it to handle other use cases
as well. 

\bd
\item [Key] Type is a struct of between 1 and 4 atomic values e..g,
\begin{verbatim}
typedef struct _keytype { 
  uint32_t key1;
  double key2;
  char key3[16];
  } keytype;
\end{verbatim}
\item [Value]
Consists of following. 
\be
\item \verb+ uint32_t count+ number of times a key can occur is \(\leq 2^{31}\) 
\item \verb+ uint32_t guid+ number of unique keys \(\leq 2^{31}\) 
\ee
\ed

\section{new}
\label{new}
To create a KeyCounter, we pass in 
\be
\item {\tt vecs} --- an indexed table of vectors, all of which should have the same length and the same chunk size
\item {\tt optargs} --- a table which allows us to over-ride default values for
\be
\item name of KeyCounter. Has no value other than for debugging
\item label of KeyCounter. If label is {\tt foo}, then we creare
\be
\item a directory {\tt foo} in \verb+~/Q/TMPL_FIXHASHMAP/KEY_COUNTER/+
\item a library {\tt libkcfoo.so} used by LuaJIT's {\tt ffi.load()}
\ee
\ee
\ee

Internal members of KeyCounter class are 
\be
\item vecs, vectors used to create it
\item is\_eor, whether all vectors have been fully consumed
\item sum\_count, number of items inserted
\item \ldots
\ee
\section{clone}
\label{clone}

When applied on a KeyCounter, returns a brand new empty KeyCounter.
Takes same arguments as Section~\ref{new} with the following differences
\be
\item The number and type of the table of {\tt vecs} must match the ones used to
create the original KeyCounter
\item can set name in optargs but not label
\ee

\section{next}
\label{next}
Consumes one chunk from each of the vectors passed in.
{\tt is\_eor} is set to true when nothing more to consume.

\section{eval}
\label{eval}
Calls Section~\ref{next} repeatedly until all vectors have reached {\tt eov}.
Note that all vectors must have same length.

\section{label}
\label{label}
Returns label of KeyCounter. This cannot be modified. 
Set in Section~\ref{new}

\section{name}
\label{name}
Returns name of KeyCounter. 

\section{set\_name}
\label{set_name}
If no argument, then name set to nil.
Else, must pass string which is used as new name

\section{size}
\label{size}
Returns size of hash table

\section{nitems}
\label{nitems}
Returns number of items in hash table. Always less than size.

\section{is\_eor}
\label{is_eor}
Returns {\tt is\_eor} --- {\tt true} when vectors have been fully consumed; {\tt
false} otherwise.

\section{sum\_count}
\label{sum_count}
Returns the number of insertions into the hashmap.

\section{get\_val}
\label{get_val}

Given an indexed table of scalars that represents a key, 
returns the key and value and position in the hash table where that key
occurred. 

\be
\item Number of scalars same as number of vectors used in Section~\ref{new}
\item Types of scalars must match types of vectors used in Section~\ref{new}
\item Special case when type is {\tt SC}. The length of the string must be
lesser than the width of the corresponding vector.
\ee

Returns are 
\be
\item {\tt key}, pointer to the key struct inside the hash table. 
\item {\tt keytype}, string that can be used to cast returned key to proper tyope
for access from Lua 
\item {\tt val}, pointer to the value struct inside the hash table. 
\item {\tt valtype}, string that can be used to cast returned val to proper tyope
for access from Lua 
\item {\tt bool is\_found}, indicating whether the key was found
\item {\tt uint32\_t where\_found}, which is where the key was  found
in the hash table. 
\ee

\section{condense}
\label{condense}
Input is a field which can be either 
\be
\item {\bf native} field. Return type is I4, ideally UI4. Options are 
\be
\item count (non-zero)
\item guid (non-zero)
\item idx
\ee
\item {\bf auxiliary} field . Options are 
\be
\item cum\_count. Type is I8, ideally UI8.
\item \ldots
\ee
\ee
Output is a vector. Creation of this vector is 
best explained with an example. Let us say that the hash table looks like
Table~\ref{hash_tbl_1}. Depending on the input, we get back one of the columns
of Table~\ref{condensed_hash_tbl_1}.
\begin{table}
\centering
\begin{tabular}{|l|l||l|l|l||l|} \hline \hline 
{\bf Key1 } & {\bf Key 2} & {\bf count} & 
{\bf guid} & {\bf idx} & {\bf cum\_count} \\ \hline \hline
--- & --- & --- & --- &  0 & 0  \\ \hline
  1 &  6  &  3  &  1  &  1 & 0  \\ \hline
  2 &  5  &  2  &  2  &  2 & 3  \\ \hline
--- & --- & --- & --- &  3 & 5  \\ \hline
  3 &  4  &  1  &  3  &  4 & 5  \\ \hline
  4 &  3  &  6  &  6  &  5 & 6  \\ \hline
  5 &  2  &  5  &  5  &  6 & 12  \\ \hline
--- & --- & --- & --- &  8 & 17  \\ \hline
  6 &  1  &  4  &  4  &  8 & 17  \\ \hline
\hline
\end{tabular}
\caption{Sample Hash Table}
\label{hash_tbl_1}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l|l||l|} \hline \hline 
count & guid & idx & cum\_count \\ \hline \hline
  3  &  1  &  1 & 0 \\ \hline
  2  &  2  &  2 & 3 \\ \hline
  1  &  3  &  4 & 5 \\ \hline
  6  &  6  &  5 & 6 \\ \hline
  5  &  5  &  6 & 12 \\ \hline
  4  &  4  &  8 & 17 \\ \hline
\hline
\end{tabular}
\caption{Condensation of Table~\ref{hash_tbl_1}}
\label{condensed_hash_tbl_1}
\end{table}

\section{make\_cum\_count}
\label{make_cum_count}

Makes an auxiliary field alongside the native fields {\tt count}
and {\tt guid}. This is creates from {\tt count}, example in 
Table~\ref{condensed_hash_tbl_1}.

\section{make\_permutation}
\label{make_permutation}
Input is an indexed table of vectors which must match the vectors used to create
the KeyCounter. Output is a vector of type I4, ideally UI4.

Best explained with an example. 
Consider hash table in Table~\ref{hash_tbl_2} with 4 items and a 
cumulative count of 10. Assume the input to Section~\ref{make_permutation} is in
Table~\ref{data_1}. 

Note that the guarantee about the output vector is 
\be
\item it is a permutation on \(0, 1, 2 \ldots n-1\) where \(n\) is the length of the input vectors.
\item the values for a given key are contiguous. For example, 
\be
\item values for key \((A, 1)\) are \((3, 4, 5, 6)\)
\item values for key \((B, 2)\) are \((7, 8, 9)\)
\ee
\ee
\begin{table}
\centering
\begin{tabular}{|l|l||l|l|l||l|} \hline \hline 
{\bf Key1 } & {\bf Key 2} & {\bf count} & {\bf guid} \\ \hline \hline
--- & --- & --- & ---  \\ \hline
  D &  4  &  1  &  4   \\ \hline
--- & --- & --- & ---  \\ \hline
  C &  3  &  2  &  3   \\ \hline
  A &  1  &  4  &  1   \\ \hline
  B &  2  &  3  &  2   \\ \hline
--- & --- & --- & ---  \\ \hline
\hline
\end{tabular}
\caption{Sample Hash Table --- 2}
\label{hash_tbl_2}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|l||l|l|l|} \hline \hline 
{\bf Key1 } & {\bf Key 2} & {\bf Permutation} & {\bf hidx} 
& {\bf count} \\ \hline \hline
A & 1 & 3 & 69  & 4 \\ \hline
B & 2 & 7 & 538 & 3 \\ \hline
C & 3 & 1 & 133 & 2 \\ \hline
D & 4 & 0 & 142 & 1 \\ \hline
A & 1 & 4 & 69  & 4 \\ \hline
B & 2 & 8 & 538  & 3 \\ \hline
C & 3 & 2 & 133 & 2 \\ \hline
A & 1 & 5 & 69 & 4 \\ \hline
B & 2 & 9 & 538 & 3 \\ \hline
A & 1 & 6 & 69  & 4 \\ \hline
\hline
\end{tabular}
\caption{Data For Permutation}
\label{data_1}
\end{table}

\section{get\_hidx}
\label{get_hidx}

Structurally similar to Section~\ref{make_permutation}. The difference is that
the output vector returns the position of the key in the hash table. See
Table~\ref{data_1}.

Conditions on the output vector are 
\be
\item Type is UI4
\item If the \(i^{\mathrm{th}}\) composite key is the same as the 
\(j^{\mathrm{th}}\) composite key, then the 
\(i^{\mathrm{th}}\) row of the output vector is the same as the 
\(j^{\mathrm{th}}\)  row of the output vector.
\item minimum value is 0, maximum value is \(n-1\) where the hashmap has \(n\)
entries.
\ee

\section{map\_out}
\label{map_out}
Input is 2 vectors
\be
\item {\tt hidx}, created by Section~\ref{get_hidx}
\item {\tt name}, a string containing name of field to be mapped out. 
This can be 
\be
\item a native field . Options are 
\be
\item {\tt count}
\item {\tt guid}
\ee
\item an auxiliary field created as in Section~\ref{make_cum_count}
\ee
\ee
Output is a vector such that 
\be
\item type is that of the field to be mapped out.
\item length is the same as that of input vector {\tt hidx} 
\ee
As an example, column with header {\bf count} in Table~\ref{data_1} 
shows what the output vector look like if the field was ``count'' for the state
of the hash table as described in Table~\ref{data_1}.

If a row of the hidx vector points is invalid (empty bucket or greater than size
of hashmap), then the corresponding value of the output vector is 0.

