\begin{abstract}
The unspoken secret of ``big data'' is that while data warehouses are
exploding in size, the actual amount of data needed for many analytics and
machine learning problems is better characterized as ``medium data''. We define
``medium data'' as data large enough that one must be cognizant of performance but small
enough that a single machine is adequate for interactive usage. This realization has motivated the
design and development of \Q, an analytical database. It has its own language
with an extensible set of operators.  The language is inspired by
Ousterhout's dichotomy, using a scripting language (Lua) with a fast interpreter
(LuaJIT) as the front-end and a compiled language (C) for the back-end.
A fundamental realization that has influenced \Q's design is that data
practitioners often build up tremendous intuition about the data sets that they
are working with. \Q\ encourages the use of this intuition to gain performance
at relatively low complexity.
\Q\ exposes its software architecture in a disciplined and layered fashion. By
this we mean that there is a continuum of optimizations that become possible as
one is willing to engage with more of the details of the system. 
The other central requirement guiding Q's development is ease of development and
interactive development, because most data science activities involve trial and error, changing requirements, etc.
To support this, we have an extreme focus on performance and 
have built a Python interface so that users can use their
favorite IDEs as well as Jupyter notebooks.

\end{abstract}
